\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage[onehalfspacing]{setspace}
\usepackage[top=2.5cm, bottom=2.5cm, left=2.5cm, right=2.5cm]{geometry}
\usepackage{mathptmx}
\usepackage{enumitem}
\usepackage[style=ieee]{biblatex}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{dirtree}

\hypersetup{
      colorlinks=true,
}
\setlist{nosep}

\addbibresource{bibliography.bib}

\title{A functional programming language\\Maturaarbeit}
\author{Enea Giger\\Betreuungsperson: Adrian Lüthi}
\date{Gymnasium Burgdorf\\\today}

\newcommand{\importListing}[1]{
    \begin{minipage}{\linewidth}
    \input{#1}
    \end{minipage}
}

\begin{document}
\maketitle
\newpage

\begin{abstract}
	This paper describes the creation of a new functional programming language
	with both an interpreter and a compiler called Imp from scratch.
	The language is aimed to be relatively simple, so it could be used as a
	teaching tool.
	It describes the essential architecture and algorithms used for those
	implementations.
	The language is written almost exclusively in the relatively new language
	Zig.
\end{abstract}

\tableofcontents
\newpage

\section{Introduction}
In todays world, computers are omnipresent. Everything depends on them.
And all these computers have to be programmed to do something useful.
Those programs need to be written in some language, which today most
often are languages like C++, JavaScript or Python.
But with recent developments, it seems that parallelism is getting more
and more important. Nowadays, most of the progress in CPUs stems not from
improved performance in a core, but from increasing the number of cores included.
Yet in most languages used to program nowadays, parallelism is usually hard to introduce.

Functional programming languages solve this problem by eliminating side effects
like mutation of variables in a function, which prevent functions to be run in
parallel, since the output of one function might depend on the variable being mutated.
This means that functional languages could be easier to parallelize and therefore
become very important in the future.

The main goal of this project was therefore to devise a functional language
and create an implementation of it.
The principal aim for the syntax of the language was then to keep it simple,
so it could be used to show and explain the concept of functional languages.
Another aim was to create this implementation with as few dependencies as possible
in order to learn all the basics from scratch.

This language was to be called Imp for Impure, as Input/Output is allowed in this language
without distinguishing functions using it from functions that do not.

\section{Fundamentals}
Though programming languages can be very different,
the elementary steps taken in each implementation are
still mostly the same.

These next paragraphs explain the basic knowledge needed
to understand the structure of this project.

\subsection{The Lexer}
The first step of a programming language implementation is almost always the lexer.
It is responsible for splitting up the source code into tokens, little chunks like the name of a variable or a statement.
This can most often be done in one single loop over the code.
In the implementation of Imp, this is the case.
These tokens can then be used as input for the parser.

\subsection{The Parser}
The Parser of a programming language is responsible for ensuring the syntax is correct and, in most cases,
building an Abstract Syntax Tree (AST) out of the tokens.
This can be done in many ways, though what methods are applicable
depend on the syntax of the language.
\subsubsection{Pratt Parsing}
Pratt Parsing \autocite{prattTopOperatorPrecedence1973}
is the method that is used for the expressions in this language.
It works with a list of rules for what to do when a certain type of token is encountered
either at the start or in the middle of an expression.
When parsing an expression using this method,
it starts with the right-binding power of 0.

Then the first type of token is looked up
in the list of rules for the start
(in the null denotation, as it's called in the paper).
This results in a function which is then called
which parses the expression starting with this token.
For example, if this is a negation, it also parses the argument
by recursively calling the expression function again.
Once this prefix is parsed the algorithm enters a loop:
\begin{enumerate}
	\item \label{start}
	      Lookup the next token in the in- and postfix table.
	      (This is called the left denotation in the paper.)
	      This results in a left-binding power and a function
	\item If there is no function listed or
	      the left-binding power stored is less or equal
	      to the right-binding power, return the current expression
	\item Otherwise, call the parsing function listed with the current expression
	      and store the result back in the current expression
	\item Return to step \ref{start}.
\end{enumerate}
For infix operators like multiplication, this parsing function
will call the expression function again, but with a higher right-binding power
so as to ensure for example that addition can't be parsed in a product,
since the left binding power of addition is less than
the right-binding power of multiplication,
which is the way precedence is solved by this algorithm.
By changing the numbers, it can also be regulated whether the operator
is calculated left-to-right or right-to-left.

\subsection{Type Checking/Inference}
After the program has been parsed, there traditionally follows
a semantic analysis step. This step most often consists of a type checker
(if the language is typed) and an optimizing step.

The Type Checking step ensures that all programs are well-typed,
meaning only operations are applied on the arguments, that can
be done, so no runtime type error can occur as it does in dynamically typed
languages like Python.

\subsubsection{Hindley-Milner type inference}
The Hindley-Milner type inference, as described in
\autocite{damasPrincipalTypeschemesFunctional1982}
is originally a way to determine types for functions in the
polymorphic lambda calculus with an addition of let expressions.

\paragraph{Polymorphic Lambda Calculus}
The Polymorphic Lambda Calculus is a simple model for computation.
The only operations that are possible are:
\begin{itemize}
	\item $\lambda x.\:e$,
	      which creates a function where $x$ is the name of the argument
	      and $e$ is an expression that is given as a result
	\item $x$, where $x$ is the name of a variable
	      and the result is the value of that variable
	\item $e_1\:e_2$, where both $e_1$ and $e_2$ are expressions and the result
	      is the result of the call of function $e_1$ with argument $e_2$\item $c$, where $c$ is a constant
	\item $let\:x = e_1\:in\:e_2$, where x is the name of the variable
	      which is defined to be the result of expression $e_1$ in the scope of expression $e_2$\end{itemize}

\subsubsection{The Inference Algorithm}
The Hindley-Milner type inference algorithm computes the types of expressions
by essentially first assuming a maximally generic type and refining it step by step
using information like the operations used (e.g. function calls).
If there turns out to be no type possible, the program is not well-typed.
Important for the type system to work is the definition that only let-defined variables
can be polymorphic (i.e. can take on multiple different types).

\subsection{The Optimizer}
The next step in the semantic analysis step is in most languages an optimizing step.
This is done in many passes over the code in most languages. The type of optimizations
possible depend on how the program is executed, whether the language is typed,
sometimes also depending on the target machine and many other aspects.

\subsection{The Interpreter}
The interpreter is the next step in many programming languages.
It is a program intended to run the user-written program in the new
language. There are many types of interpreters:
\begin{itemize}
	\item \textbf{Byte-Code Interpreters}, which run a
	      a custom-defined intermediate representation that is similar to how
	      machine code works, but more specific to the language. Most interpreted languages
	      use one, because it is most often the fastest possibility.
	\item \textbf{Threaded Interpreters}, where there is no byte sequence, but pointers
	      to functions to be called or the next instruction. This has the problem that there
	      are many indirect calls and pointer indirections being done, reducing the speed.
	\item \textbf{Abstract-Syntax-Tree Interpreters}, which directly evaluate the tree,
	      which has more information available, but might not be optimal because of syntax
	      nodes which need to be skipped and there are more pointer indirections than
	      \textbf{Byte-Code Interpreters}
\end{itemize}
Both the byte-code and threaded interpreters require a compiler.

\subsection{The Compiler}
The Compiler uses the parsed program and converts it into another langugage,
most often the machine language. Programs using the machine language are most often
faster than ones run by interpreted languages, thanks to directly being able to use
optimizations in the CPU, which can be used since the program is specific
and not general like an interpreter.

Many compilers nowadays don't directly convert the code to machine code,
but depend on a backend. One of the most popular backends is \textbf{LLVM}.

\subsubsection{LLVM}
\textbf{LLVM}, which originally stood for Low Level Virtual Machine, is
a backend capable of converting a custom language called
LLVM IR (Intermediate Representation) into code for several targets,
like x86-based cpus (most Intel and AMD chips), ARM-based cpus, but also
WASM (Web Assembly, a language understood by browsers).

It is coded in C++, though bindings exist for several languages, most notably
for C, which many languages can interact with.

\section{Implementation}

\subsection{Used Software}

The Software was developed using the Editors Visual Studio Code and Zed.
Zed was used primarily, with VSCode used for features like Debugging
that didn't work fully in Zed.

The Programming Language itself was implemented in Zig, a new low-level programming
language that aims to improve on C. Zig was used in this project, since it has builtin
support for tagged unions and a build system. Tagged Unions were deemed important because they
enhance type safety for things like the AST because of the different node types.
However, for the compiler, C was still used for
a few builtin functions.

LLVM was used as a tool to compile Imp to machine code while the
Boehm-Demers-Weiser garbage collector \autocite{GarbageCollector} was
used to manage memory in the compiler.

For the ability to syntax highlight code in Imp, an additional parser was made using tree-sitter.
This parser is written in JavaScript. Since the grammar of the language is too complex,
a custom lexer was also needed, which itself was written in C.
To state what parts of the code should be colored how, Queries had to be written as well,
which are written in an unnamed language unique to Tree-sitter.

Highlighting with Tree-sitter was chosen specifically since it is used
in the editor Zed, enabling a custom extension to be written for Imp.

This article itself was written in LaTex, while the listings were generated
by chromacode\autocite{lebedaTomLebedaChroma_code2025}.


\subsection{The Language}
\importListing{code/either.tex}
\importListing{code/foldl.tex}

The Syntax of the language consists of statements and expressions.

There are only two types of statements.
\begin{itemize}
	\item Type statements, which define an algebraic data type,
	      meaning a type whose values can be constructed by various
	      constructors when given specific types.
	      These types can be generic.
	      An example can be seen in Listing \ref{listing:either}, where the type Either is defined,
	      which can be constructed by either a value of type a or of type b.
	\item Let statements, which define an (immutable) variable
	      in the rest of the file. If the variable is assigned a
	      lambda expression, it can be recursive
\end{itemize}
Statements are separated from expressions by not being indented,
which has the consequence that all expressions have to be indented,
but makes it so that there does not need to be an end keyword
for these statements.
This is the only case of significant whitespace in the language.

The types of expressions that exist are:
\begin{itemize}
	\item Lambda expressions, which work exactly
	      the same as in pure lambda calculus
	\item Let expressions, which work the same as in pure
	      lambda calculus, except, as with the statement,
	      it can be recursive if assigned a lambda
	\item Variables,
	\item Function Calls,
	\item Constants, meaning Integers (whole numbers), Floats,
	      One-Byte Unicode Characters, and Booleans,
	\item If expressions,
	\item Operator applications (both prefix and infix),
	\item Case expressions, which match a value of an ADT
	      based on the Constructor,
	\item and syntax sugar for lists and strings,
	      which are lists of characters
\end{itemize}

Comments start with a $\#$ sign and end with the end of the line.

An example of code in this language is the fold left function
given in Listing \ref{listing:foldl}.

\subsection{The Lexer}
The lexer of Imp is a simple single-pass lexer, that computes
the tokens on-demand. This is done to avoid additional memory
allocations other than creating the AST.

\subsubsection{Token Types}
\begin{itemize}
	\item The Keywords $let$ and $in$, $lambda$, $if$, $then$ and $else$,
	      $case$ and $of$, $or$, and $and$,
	\item The special characters $.$, $=$, $:$, $($ and $)$, $[$ and $]$
	\item The special sequences $->$ and $=>$,
	\item all operators for comparison and mathematical operations,
	      and the special operator $;$,
	\item Numbers, Booleans, Characters and Strings,
	\item Identifiers,
	\item and Special Tokens for new lines and the end of the file
\end{itemize}

\subsection{The Parser}
\importListing{code/expressions.tex}

The Parser consists of a pretty straight-forward implementation of the Pratt Parsing algorithm
for the expressions as shown in Listing \ref{listing:expressions} and a simple loop for statements.
In this implementation, the lookup table is represented by two functions,
$getNud$ and $getLed$.
The problem that there might not be an entry is solved by returning null
in that case.

\subsection{Type Checking/Inference}
The Type Checker is also mostly an implementation close to the Algorithm J as described
in the paper \autocite{damasPrincipalTypeschemesFunctional1982}
There is however a significant difference to the original type system:
There exists a special type Number(a), which can be either a Float or an Int
and therefore can be unified by them Both. When unified, the runtime representation
is then Number(Int) or Number(Float).
It is printed similarly to how Type Class constraints are in Haskell.
This means that the type Number(a) is printed as \texttt{Number a => a}.
However, in order to be able to compute free type variables for generalization
faster, levels as described in \autocite{EfficientInsightfulGeneralization2022} are
used. However, they are called depth in the code.

\subsection{The Optimizer}
\importListing{code/optimizer.tex}

The Optimizer works slightly different depending on whether
the code is interpreted or compiled. In the case that it is interpreted
the optimizer also combines lambdas into larger objects,
while the compiler doesn't need this. However, in both cases
a list of bound variables is gathered per function definition,
since all functions are closures.
The outline of the optimizer is shown in Listing \ref{listing:optimizer}.

\subsection{The Interpreter}
The Interpreter is a simple tree-walking Interpreter,
meaning it directly uses the AST to run the code.
The most important optimization it employs is proper tail call elimination,
meaning it doesn't make another stack frame if the function call is the last
thing a function does, making recursion feasible through
tail recursion. This is done by creating a simple form of trampolining,
where the evaluation of an expression returns the next thing to
do, which is then repeatedly evaluated in a loop until the final result
is reached.

\subsection{The Monomorphizer}
The Monomorphizer is a step only needed for the compiler.
It essentially eliminates all polymorphism by duplicating
the value for each type.

This is done by a simple pass over the AST, gathering
all the types a value takes on in the body of each let expression
or in the rest of the file and then copies the AST of the definition once for each
type.

This method is essentially the same as MLTon uses.\autocite{Monomorphise}

\subsection{The Compiler}
The Compiler converts then converts this
monomorphic AST into LLVM IR using the C API of LLVM through
Zigs automatically generated FFI.

Closures are represented in the IR as a struct consisting of two pointers:
The first being the function pointer, the second being a pointer to the heap
where all bound variables are stored.

The Functions itself take two arguments each: Their actual argument
as declared by the source code, and the environment that had been stored
at the time of the closures creation.

Here, little is done to optimize the IR created apart
from recursive calls being done directly instead of through an indirect
call, as all other function calls are done in order to enable
the tail recursion optimization as it could not resolve the indirect function call
as a recursive call since it would be given via an argument and therefore
could theoretically be different.

Then clang is called on the IR, linking it with a very small library of builtin
functions with the highest optimization level of -O3. The tool therefore requires
clang to be present and on the PATH.

\section{Results}
Since the language Imp is based on the typed lambda-calculus with recursion,
it is definitely turing-complete, meaning it can compute everything any other
language could. For example, it can be used to generate the fibonacci sequence,
as shown in Listing \ref{listing:fibonacci}.

\importListing{code/fibonacci.tex}

In this example, the fibonacci sequence is computed in the function
$fibonacciHelper$ through tail recursion. Because of that, that function
can be viewed as a loop which changes the variables $previous$, $current$ and
$n$ on each iteration. This change is done on line 6.

Throughout its execution, it also prints the current fibonacci number
computed using the two builtin functions $showInt$ and $print$.
$showInt$ is a function that converts an integer into a list of characters.

Several expressions are separated by a semicolon ($;$) which is an operator
in this language that simply discards the result on the left, meaning
the expression on the left is only there for its side-effects like printing.

The main function takes a dummy argument x, because in Imp, there are no
functions that take no argument. This function is then called
after all definitions have been executed.
On line 17 it then reads the number of elements requested,
which is then parsed. As this can fail, this returns a
value of type Option. The case expression then evaluates either line
18, if the parse failed, which just repeats the main function,
or it puts the result into the variable n which is then used in
lines 20-23. Finally on line 24, Void is returned, since Imp functions
also need to return exactly one value.

This example also demonstrates the power of Hindley-Milner type inference,
because no types did ever need to be annotated for the code to work,
but it is still garanteed that the whole program has no type errors.
The type given to fibonacciHelper is only there for clarity.

Other examples of Imp code can be found in the folder examples of
in the source code.

To get a view of how the LLVM IR transformation might work, here is
an example of the resulting IR of a simple function,
though cleaned up to remove clutter and add meaningful names.

\begin{minipage}{\linewidth}
	\input{code/addImp.tex}
	\input{code/addLLVM.tex}
\end{minipage}

So the first lambda defined on line 2 of Listing \ref{listing:addImp} got transformed
into the function $@add$ of Listing \ref{listing:addLLVM}.
As the second lambda binds the value of $x$, $@add$ allocates
enough memory to store it on lines 3-5, then gets the exact location on line 6, which
is done since there might be more than one bound variable, but in this case is redundant.
Line 7 then stores $x$ while line 8 creates the closure by inserting the pointer to the bound
variable in a struct.
Finally on line 9, the closure is returned.

The function $@add_const$ then acesses this bound variable, again first getting the
location on the redundant line 14, then loads the actual value of $x$ on line 15
and finally computes the result on line 16 which is returned on line 17.

This IR will then get optimized in the final compilation by clang,
oftentimes eliminating the indirect function call through the closure.
\section{Discussion}
As seen in the example given, the language is fully functional and
can be used for mathematical computation. However, there are also several
shortcomings that will be mentioned here.

\subsection{Shortcomings}\label{shortcomings}
The language Imp is a functioning prototype, but there are several
aspects which would need to be treated in order for this language
to be useful for any practical purposes.

\subsubsection{Input/output}
Input/output is an essential part of any programming language.
The language Imp is no exception to this, having the two functions
$print$ and $read$ which print to standard output and read from
standard input respectively. However, there is currently no way
to read from or write to files, which is essential for a program
to store information for later usages of it or to load and/or output larger
amounts of information.

To enable this feature, a type for Files or Buffers would have to be
introduced for fast, type-safe interaction.

\subsubsection{Library System}
In Imp, it is impossible to import other files from a program.
This would be essential however for the development of libraries.
To enable imports and therefore libraries, a topological sorting
algorithm could be used, since the order of definitions does
matter in this language and circular references can't exist anyway.

This would enable a much more practical implementation of the standard
library as well, since e.g. functions that operate on functions could be used
in almost all other parts of the library.

\subsubsection{FFI}
At the moment, Imp cannot access system libraries or libraries
written in other languages at all, meaning GUIs or even most TUIs
(terminal user interfaces) are impossible to create,
since you couldn't use libraries like QT.

This could be mitigated by a large set of builtin functions based
on which a standard library could be built, or more simply, though
less securely, using an FFI (foreign function interface).

Using a foreign function interface, one could declare functions that
exist in another language (most likely C), and their types and then use
those functions. This would lead to other challenges however, since
there exists no equivalent to many of the types in C.
The types would however need to match up exactly,
putting a huge burden on both the programmer writing the declarations
as there should be no type errors, otherwise the whole type system
would be for nothing, and the compiler, as it would no longer
be free to represent its objects however it wants.
LLVM does also not garantee correct FFI, even if the types
seem to match up.

\subsubsection{OS Compatibility}
As the Compiler of this language depends on both
LLVM and Boehm GC, it can only be used on Operating Systems
that support both of these. However, Linux, MacOS and Windows
are all supported by them both.

Windows has however the problem of there not being
a prebuilt package for LLVM which includes the headers as well
as the libraries to be linked. This means that any
user on windows wanting to try the compiler
must first build LLVM for himself at the moment.

\subsection{Room for improvement}
There are many parts of both the compiler and the
interpreter, which could be improved significantly
for enhanced speed and better memory consumption.

\subsubsection{Garbage Collector}
Though both the compiler and the interpreter produce
garbage collected programs, the compiler uses the
Boehm-Demers-Weiser GC \autocite{GarbageCollector},
which is unaware of any of the types of Imp and therefore
has to assume that anything that could be a pointer,
is in fact a pointer.
Meanwhile, the interpreter uses a very basic mark-and-sweep
Garbage Collector. However this could very easily be improved
by using methods like Generational Garbage Collection, where
young objects are checked more often than older ones.

\subsubsection{Defunctionalization}
Furthermore, another possible optimization done by almost all
functional programming languages nowadays, is some form of
defunctionalization, which is a process which tries
to eliminate functions as arguments, replacing them with
an object that is given as an argument which denotes which
function it represents.

One new interesting approach that seems to be promising
is the technique of lambda set specialization
\autocite{brandonBetterDefunctionalizationLambda2023a}, which
could be applied to this language relatively easily,
since it assumes a Hindley-Milner style type inference
as well.

\section{Conclusions}
In conclusion, all the goals were met and the project was a success.
Though there were a few roadblocks, especially concerning LLVM,
the language is now fully functional and able to be tested by anyone.
There are a few aspects that would need to be improved for it to be
useful for anything practical as explained in section \ref{shortcomings},
it is already a fully useable language that could be used as a teaching
tool.

\newpage
\printbibliography

\lstlistoflistings

\appendix
\section{Source Code}
The source code for this project can be found at: \href{https://github.com/enmiligi/matura-project}{GitHub}

\end{document}
